---
title: "DataDeficiency_analysis"
author: "Alex"
date: "8/29/2019"
output: html_document
---

### This is an R Markdown file that will show the annotated analysis of our data on factors contributing to data deficiency in all known shark species.

```{r}
library(tidyverse);library(lme4);library(sjPlot)
# Read in data
dd <- read.csv("DD_analysis_datafile.csv", stringsAsFactors = FALSE)
dd = dd[dd$Order != "",]

# Set NAs to 0
dd[is.na(dd)] = 0

# Rename some columns
dd = dd %>% rename("Rep_Strategy" = "X.Reproductive.Strategy",
                             "Vulnerability" = "Vulnerability..",
                             "Antarctic" = "Antartic")

# Cleaning up fisheries data
dd$Fisheries[dd$Fisheries == "N/A"] <- "Unknown"
dd$Fisheries[dd$Fisheries == ""] <- "Unknown"
dd$Fisheries <- factor(dd$Fisheries, levels = c("Unknown", "0", "1"))

# Cleaning up reproductive strategy data
dd$Rep_Strategy[dd$Rep_Strategy == ""] = "Unknown"
dd$Rep_Strategy = factor(dd$Rep_Strategy, levels = c("Unknown", "Ovoviviparous", "Viviparous", "Oviparous"))

# Removing columns unused in data
dd$Benthic <- as.numeric(dd$Benthic) %>% replace_na(0)
dd$Vulnerability <- as.numeric(dd$Vulnerability) %>% replace_na(0)

## Check normality of size variable; needs to be log transformed
hist(dd$Size,100)
dd$log_size<- log(dd$Size) #added a column with log transformed values, which we will use in the GLM

summary(dd)

#simple look: 
#Out of 501 shark species assessed: 37.5% are classified as DD, with ~ 4% not evaluated
sum(dd$Data.Deficient==1, na.rm=T)/nrow(dd) #.375
sum(dd$Data.Deficient==0, na.rm=T)/nrow(dd) # .585
```

### After we have our superficial values describing the data, time to do some stats and visualization. What potential predictive factors are we interested in?

*Biology*  
Size  
Reproductive strategy  

*Human Use*  
Vulnerability  
Type of fishery (commercial, gamefish, aquaculture)  

*Ecology*  
Geographic range  
Latitude/temperature (tropical/temperate)  
Habitat (benthic, deepwater, pelagic, coastal, freshwater)  
Depth range  


```{r}

#First, we will look at the coefficients of every single first order variable and extract p-values to see what we are dealing with
firstorder.mod <- glm(Data.Deficient ~ Deepwater + Coastal + Pelagic + Benthic + BrackishFreshwater +
      Tropical + Temperate + 
      Global + Pacific + Atlantic + Indian + Arctic + Antarctic + Trans.oceanic +
      Fisheries + commercial + subsistence + Gamefish + Aquarium + Vulnerability + 
      log_size + Rep_Strategy, data = dd, family = "binomial")

firstorder.mod 

### note that we did try to include "Family" as a random variable; however, it was "rank deficient", which means that it was correlated with multiple other variables and therefore redundant. However, code is below

firstorder.mod.re <- glmer(Data.Deficient ~ Deepwater + Coastal + Pelagic + Benthic + BrackishFreshwater + 
      Tropical + Temperate + 
      Global + Pacific + Atlantic + Indian + Arctic + Antarctic + Trans.oceanic +
      Fisheries + commercial + subsistence + Gamefish + Aquarium + Vulnerability + 
      log_size + Rep_Strategy +
        (1|Order), data = dd, family = "binomial")


```
### Now to visualize a coefficient plots to see what is significant!

```{r}
# Pulll out plot data
coefs <- plot_model(firstorder.mod, transform = NULL)

# Making custom figure
coefs$data %>%
  mutate(term = factor(term, levels = term)) %>%
  # Color by sign of change, transparency by p-value
  ggplot(aes(x = estimate,
             y = as.numeric(term),
             color = as.factor(sign(estimate)),
             alpha = as.factor(p.value < 0.05))) +
  geom_point() + 
  # Add CIs around point
  geom_segment(aes(x = conf.low,
                   xend  = conf.high,
                   yend = as.numeric(term))) +
  geom_vline(xintercept = 0) +
  # Adding labelled y axis
  scale_y_continuous(labels = as.character(coefs$data$term),
                     breaks = seq(1, 23, 1)) +
  scale_alpha_discrete(range = c(.5, 1)) +
  xlab("Log Odds Ratio of Being DD (Higher Values = Greater Chance)") +
  ylab("Variables")
```

# Results  
Here we see a few significant relationships:  

*Biology*  
**Size** negative correlation with DD  
Reproductive strategy  

*Human Use*  
**Vulnerability** positive correlation with DD  
Type of fishery (commercial, gamefish, aquaculture)  


*Ecology*  
**Geographic range** positive corr. with Arctic Ocean (note small sample size); negative corr. with Indian Ocean   
Latitude/temperature (tropical/temperate)  
**Habitat** (**benthic - positive correlation**, deepwater, pelagic, coastal, **freshwater - negative correlation**)  
Depth range  


### Let's visualize our significant results

```{r}
## subset coefficients by category

fullcoeffs <- summary(firstorder.mod)$coefficients
geo_coeffs <- fullcoeffs[c("Atlantic", "Arctic", "Global", "Indian", "Pacific", "Global"),] #note that this is a matrix
bp <- barplot(geo_coeffs[,"Estimate"], ylab="Model coefficients")

size_coeffs <- fullcoeffs[c("log_size"),]
size_coeffs
### Plot backtransformed values against all values of size in this data set (dd$Size vs. )

vuln_coeffs <- fullcoeffs[c("Vulnerability"),]
### See same thing for size, except with categorical variables

hab_coeffs <- fullcoeffs[c("Benthic", "Deepwater", "Pelagic", "Coastal", "BrackishFreshwater"),]
bp2 <- barplot(hab_coeffs[,"Estimate"], ylab="Model coefficients")
```

